Termination Analysis of Imperative Programs Using Bitvector Arithmetic

### 2016-09-09 paper-discussion/main.tex
renamed the directory for paper discussion slides

\begin{columns}
  \begin{column}{4.5cm}
\item Terminating wrt bitvector arithmetic, but non-terminating wrt
  integer arithmetic.

\begin{exampleblock}{Example}
\begin{verbatim}
void f(int i) {
  while (i > 0)  
    i++;
}
\end{verbatim}
\end{exampleblock}

@@

\end{column}
\end{columns}

@@

\begin{block}{Bird's-eye view of \texttt{KITTeL}'s approach}
\begin{figure}
  \includegraphics[width=\textwidth]{view.png}
\end{figure}
\end{block}

@@

\begin{frame}{From C to LLVM-IR}
From C code to LLVM-IR code, and then \emph{basic block graph}:
\begin{exampleblock}{Example 1}
\begin{figure}
  \includegraphics[width=\textwidth]{CtoLLVM.png}
\end{figure}
\end{exampleblock}
\end{frame}

\begin{frame}{From LLVM-IR to Transition System}
The corresponding transition system of Example~1:
\begin{exampleblock}{Example 1}
\begin{figure}
  \includegraphics[width=\textwidth]{LLVMtoTS.png}
\end{figure}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{From Transition Systems to TRSs}
\begin{itemize}
\item A transition system can be represented in the form of an
  \texttt{int}-based TRS.
\item Each transition $s_1 \rightarrow^\lambda s_2$ gives rise to a
  rewrite rule
\begin{center}
$s_1(x_1,\ldots,x_n) \rightarrow s_2(e_1,\ldots,e_n) ~ \llbracket\varphi\rrbracket$
\end{center}
\end{itemize}
\begin{exampleblock}{TRS of Example 1}
\footnotesize
\begin{tabular}{rcl}
$\verb|entry|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|bb1|(\verb|x|, \verb|y|, \verb|y|, 1, \verb|%1|, \verb|%2|)$ \\
$\verb|bb1|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|bb|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|) 
  ~ \llbracket \verb|y.0| > 0 \rrbracket$ \\
$\verb|bb1|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|return|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|) 
  ~ \llbracket \neg(\verb|y.0| > 0) \rrbracket$ \\
$\verb|bb|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|bb1|(\verb|x|, \verb|y|, \verb|y.0| - 1, \verb|r.0|, \verb|r.0| * \verb|x|, \verb|y.0| - 1)$
\end{tabular}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Reducing TRSs}
The number of arguments, hence the TRS, can be further reduced by
standard compiler techniques:
\begin{itemize}
\item \emph{Backward slicing} removes all variables that are not
  relevant for the control flow of the program. (\verb|x|, \verb|r.0|,
  \verb|%1|, and \verb|%2| in Example~1)
\item \emph{Liveness analysis} removes variables before they are
  defined or after they are no longer needed. (\verb|y.0| in
  \verb|entry| and \verb|return|; \verb|y| in all function symbols but
  \verb|entry|)
\end{itemize}

\begin{exampleblock}{Reduced TRS of Example 1}
\footnotesize
\begin{tabular}{rcl}
$\verb|entry|(\verb|y|)$ 
  & $\rightarrow$ 
  & $\verb|bb1|(\verb|y|)$ \\
$\verb|bb1|(\verb|y.0|)$ 
  & $\rightarrow$ 
  & $\verb|bb|(\verb|y.0|) 
  ~ \llbracket \verb|y.0| > 0 \rrbracket$ \\
$\verb|bb1|(\verb|y.0|)$ 
  & $\rightarrow$ 
  & $\verb|return|() 
  ~ \llbracket \neg(\verb|y.0| > 0) \rrbracket$ \\
$\verb|bb|(\verb|y.0|)$ 
  & $\rightarrow$ 
  & $\verb|bb1|(\verb|y.0| - 1)$
\end{tabular}
\end{exampleblock}
\end{frame}

@@

\begin{frame}{Overview of the Encoding}
The bitvector $b_{n-1}\cdots b_0$ is represented by its \emph{signed}
(two's complement) integer 
\begin{center}
$-b_{n-1}\cdot 2^{n-1} + \Sigma^{n-2}_{i=0}b_i\cdot 2^i \;\;\in \{
  -2^{n-1},\ldots,2^{n-1}-1\}$.
\end{center}

The semantics of the bitvector operations and the wrap-around behavior
of bitvector arithmetic is handled by two phases:
\begin{enumerate}
\item The conversion of \emph{comparison}, \emph{arithmetical},
  \emph{bitwise} and \emph{extension/truncation} instructions is
  adapted.  In particular, signed and unsigned operations are
  distinguished.
\item The wrap-around behavior of the arithmetical instructions is
  modeled in order to ensure that they are within the appropriate
  ranges.
\end{enumerate}
\end{frame}

@@

\medskip
The comparison instructions:
\begin{figure}
  \includegraphics[width=\textwidth]{ins1.png}
\end{figure}

For example,
\footnotesize
\begin{eqnarray}
ugt(x,y) & = & (x \ge 0 \land y \ge 0 \land x > y) \nonumber\\
& & \lor\; (x \ge 0 \land y < 0) \nonumber\\
& & \lor\; (x < 0 \land y < 0 \land x > y) \nonumber
\end{eqnarray}
\end{frame}

\begin{frame}{Phase 1 (cont'd)}
\begin{figure}
  \includegraphics[width=\textwidth]{ins2.png}
\end{figure}
\end{frame}

@@

For this, the wrap-around behavior is simulated by explicitly
normalizing the resulting integers to be in the appropriate ranges.

Each rewrite rule
\begin{center}
$\rho : f(x_1,\ldots,x_n) \rightarrow g(p_1,\ldots,p_m) ~ \llbracket \varphi \rrbracket$
\end{center}
should be replaced by:
\begin{figure}
  \includegraphics[width=0.6\textwidth]{rules.png}
\end{figure}

@@

\begin{frame}{Conclusion}
\begin{itemize}
\item The approach is \emph{sound}.
\begin{itemize}
\item The instructions are \emph{soundly} approximated.
\item The wrap-around behavior is correctly modeled.
\end{itemize}
\item The implementation is powerful than the related approach in
  [CKR+10], with a comparable average time.
\item The approach can be combined with other existing techniques for
  analyzing termination of transition systems/\texttt{int}-based TRSs.
\end{itemize}
\end{frame}

>>>>

  \begin{columns}

    \begin{column}{4.5cm}
\item Terminating wrt bitvector arithmetic, but non-terminating wrt
  integer arithmetic.

\begin{exampleblock}{Example}
\begin{verbatim}
void f(int i) {
  while (i > 0)  
    i++;
}
\end{verbatim}
\end{exampleblock}
  \end{column}

@@

\end{columns}

@@

\begin{block}{Bird's-eye view of \texttt{KITTeL}'s approach}
\begin{figure}
  \includegraphics[width=\textwidth]{figs/view.png}
\end{figure}
\end{block}

@@

\begin{frame}{From C to LLVM-IR}
From C code to LLVM-IR code, and then \emph{basic block graph}:
\begin{exampleblock}{Example 1}
\begin{figure}
  \includegraphics[width=\textwidth]{figs/CtoLLVM.png}
\end{figure}
\end{exampleblock}
\end{frame}

\begin{frame}{From LLVM-IR to Transition System}
The corresponding transition system of Example~1:
\begin{exampleblock}{Example 1}
\begin{figure}
  \includegraphics[width=\textwidth]{figs/LLVMtoTS.png}
\end{figure}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{From Transition Systems to TRSs}
\begin{itemize}
\item A transition system can be represented in the form of an
  \textcolor{red}{\texttt{int}-based TRS}.
\item Each transition $s_1 \rightarrow^\lambda s_2$ gives rise to a
  rewrite rule
\begin{center}
$s_1(x_1,\ldots,x_n) \rightarrow s_2(e_1,\ldots,e_n) ~ \llbracket\varphi\rrbracket$
\end{center}
\end{itemize}
\begin{exampleblock}{TRS of Example 1}
\footnotesize
\begin{tabular}{rcl}
$\verb|entry|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|bb1|(\verb|x|, \verb|y|, \verb|y|, 1, \verb|%1|, \verb|%2|)$ \\
$\verb|bb1|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|bb|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|) 
  ~ \llbracket \verb|y.0| > 0 \rrbracket$ \\
$\verb|bb1|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|return|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|) 
  ~ \llbracket \neg(\verb|y.0| > 0) \rrbracket$ \\
$\verb|bb|(\verb|x|, \verb|y|, \verb|y.0|, \verb|r.0|, \verb|%1|, \verb|%2|)$ 
  & $\!\!\!\!\rightarrow\!\!\!\!$ 
  & $\verb|bb1|(\verb|x|, \verb|y|, \verb|y.0| - 1, \verb|r.0|, \verb|r.0| * \verb|x|, \verb|y.0| - 1)$
\end{tabular}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Reducing TRSs}
The number of arguments, hence the TRS, can be further reduced by
standard compiler techniques:
\begin{itemize}
\item<2-> \textcolor{red}{Backward slicing} removes all variables that are not
  relevant for the control flow of the program. (\verb|x|, \verb|r.0|,
  \verb|%1|, and \verb|%2| in Example~1)
\item<3-> \textcolor{red}{Liveness analysis} removes variables before they are
  defined or after they are no longer needed. (\verb|y.0| in
  \verb|entry| and \verb|return|; \verb|y| in all function symbols but
  \verb|entry|)
\end{itemize}
\begin{exampleblock}{Reduced TRS of Example 1}
\footnotesize
\begin{tabular}{rcl}
$\verb|entry|(\verb|y|)$ 
  & $\rightarrow$ 
  & $\verb|bb1|(\verb|y|)$ \\
$\verb|bb1|(\verb|y.0|)$ 
  & $\rightarrow$ 
  & $\verb|bb|(\verb|y.0|) 
  ~ \llbracket \verb|y.0| > 0 \rrbracket$ \\
$\verb|bb1|(\verb|y.0|)$ 
  & $\rightarrow$ 
  & $\verb|return|() 
  ~ \llbracket \neg(\verb|y.0| > 0) \rrbracket$ \\
$\verb|bb|(\verb|y.0|)$ 
  & $\rightarrow$ 
  & $\verb|bb1|(\verb|y.0| - 1)$
\end{tabular}
\end{exampleblock}
\end{frame}

@@

\begin{frame}{Overview of the Encoding}
\uncover<2->{
The bitvector $b_{n-1}\cdots b_0$ is represented by its \emph{signed}
(two's complement) integer 
\begin{center}
$-b_{n-1}\cdot 2^{n-1} + \Sigma^{n-2}_{i=0}b_i\cdot 2^i \;\;\in \{
  -2^{n-1},\ldots,2^{n-1}-1\}$.
\end{center}
}

\uncover<3->{
The semantics of the bitvector operations and the wrap-around behavior
of bitvector arithmetic is handled by two phases:
}
\begin{enumerate}
\item<4-> The conversion of \textcolor{red}{comparison}, \textcolor{red}{arithmetical},
  \textcolor{red}{bitwise} and \textcolor{red}{extension/truncation instructions} is
  adapted.  In particular, signed and unsigned operations are
  distinguished.
\item<5-> The \textcolor{red}{wrap-around behavior} of the arithmetical instructions is
  modeled in order to ensure that they are within the appropriate
  ranges.
\end{enumerate}
\end{frame}

@@

\medskip
\uncover<2->{
The comparison instructions:
\begin{figure}
  \includegraphics[width=\textwidth]{figs/ins1.png}
\end{figure}
}

\uncover<3->{
For example,
\footnotesize
\begin{eqnarray}
ugt(x,y) & = & (x \ge 0 \land y \ge 0 \land x > y) \nonumber\\
& & \lor\; (x \ge 0 \land y < 0) \nonumber\\
& & \lor\; (x < 0 \land y < 0 \land x > y) \nonumber
\end{eqnarray}
}
\end{frame}

\begin{frame}{Phase 1 (cont'd)}
\begin{figure}
  \includegraphics[width=\textwidth]{figs/ins2.png}
\end{figure}
\end{frame}

@@

\uncover<2->{
For this, the wrap-around behavior is simulated by explicitly
normalizing the resulting integers to be in the appropriate ranges.
}

\uncover<3->{
Each rewrite rule
\begin{center}
$\rho : f(x_1,\ldots,x_n) \rightarrow g(p_1,\ldots,p_m) ~ \llbracket \varphi \rrbracket$
\end{center}}
\uncover<4->{
should be replaced by:
\begin{figure}
  \includegraphics[width=0.6\textwidth]{figs/rules.png}
\end{figure}
}

@@

\begin{frame}{Conclusion}
\begin{itemize}[<+->]
\item The approach is \textcolor{red}{sound}.
\begin{itemize}
\item The instructions are \emph{soundly} approximated.
\item The wrap-around behavior is correctly modeled.
\end{itemize}
\item The implementation is powerful than the related approach in
  [CKR+10], with a comparable average time.
\item The approach can be combined with other existing techniques for
  analyzing termination of transition systems/\texttt{int}-based TRSs.
\end{itemize}
\end{frame}
